---
title: "Code for analyzing conditionally rare taxa in soils"
author: "Rachel Kaminsky"
date: "6 July 2017"
output: html_document
---
R code for analyzing the contribution of conditionally rare taxa to prokaryotic populations (Kaminsky and Morales 2018, also Chapter 6 of my PhD thesis)

Load packages
```{r}
library(ggplot2)
library(ggthemes)
library(vegan)
library(phyloseq)
```

Set working directory and add files
```{r, cache = TRUE, cache.lazy=FALSE}
setwd("/Your_FilePath")
otu <- paste("otu.biom", sep = "")
map <- paste("map.txt", sep = "")
otu <- import_biom(otu)
map <- import_qiime_sample_data(map)
mm <- merge_phyloseq(map,otu)
```

Remove OTUs that aren't present in at least one sample, confirm all taxa > 0
```{r, cache = TRUE, cache.lazy=FALSE}
mm_rm = prune_taxa(taxa_sums(mm) > 0, mm)
any(taxa_sums(mm)== 0)
```

Divide counts by ten, remove zeros
```{r, cache = TRUE, cache.lazy=FALSE}
mm_10  = transform_sample_counts(mm, function(x) x / 10) 
mm_r  = transform_sample_counts(mm_10, round )
mm_r = prune_taxa(taxa_sums(mm_r) > 0, mm_r)
any(taxa_sums(mm_r)== 0)
```

Rename ranks
```{r, cache = TRUE, cache.lazy=FALSE}
colnames(tax_table(mm_r)) = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
```

For time-responsive CRT:

Subset site-level data (using site AV as an example, so must repeat this 23 more times)
```{r}
av=as.vector(sample_data(mm_r) $Site=="AV")
av=prune_samples(av, mm_r)
av=prune_taxa(taxa_sums(av)>0, av)
```

Make OTU table that averages reads from site replicates under each time point 
```{r, cache = TRUE, cache.lazy=FALSE}
speciesList <- tapply(sample_names(av), get_variable(av, "SampleTP"), c)
speciesPhyseq <- lapply(speciesList, prune_samples, av)
speciesOTUtable <- lapply(speciesPhyseq,otu_table)
speciesAvg <- lapply(speciesOTUtable,rowMeans)
pooledOTUtable = t(do.call(rbind,speciesAvg)) 
pooledOTUtable = data.frame(OTU=row.names(pooledOTUtable),pooledOTUtable)

write.table(pooledOTUtable, "av_tp_otu.txt", quote = FALSE, row.names = FALSE, sep = "\t")
```

Convert to biom in qiime

Import biom file, add metadata and taxonomy
```{r}
av_tp_biom <- paste("av_tp_otu.biom", sep = "")
av_tp_biom <- import_biom(av_tp_biom)
av_tp_map <- paste("MM_All_byTP_mapping_R.txt", sep = "")
av_tp_map <- import_qiime_sample_data(av_tp_map)
av_tp <- merge_phyloseq(av_tp_biom, tax_table(av), av_tp_map)
```

Round counts
```{r}
av_tp_r  = transform_sample_counts(av_tp, round)
```

Plot rank abundance curve (figure s2)
```{r}
library(dplyr)
phyloTemp = transform_sample_counts(av_tp_r, function(x) 1e+02 * x/sum(x))
clusterData = psmelt(phyloTemp)
clusterData = filter(clusterData,Abundance > 0)
clusterAgg = aggregate(Abundance ~ OTU + Phylum,data=clusterData,mean)
clusterAgg = clusterAgg[order(-clusterAgg$Abundance),]
ggplot(clusterAgg,aes(x=reorder(OTU,-Abundance),y=Abundance)) +
    geom_point(aes(color=Phylum),size=3) + 
    theme(axis.ticks = element_blank(), axis.text.x = element_blank()) +
    scale_y_log10()
```

Use rank abundance curve to choose a cutoff for the rare biosphere

Get CRT
```{r}
av_tp_r_dat <- data.frame(otu_table(av_tp_r))
write.table(av_tp_r_dat, "av_tp_r_otu.txt", quote = FALSE, sep = "\t")

library(TSA)
source("CRT_Functions_v1.1.R")
SimpleRareToPrev.f(otu_fp="av_tp_r_otu.txt", abund_thresh=0.0001, abund_thresh_ALL=FALSE, b_thresh=0.90, rdp_lastcol=FALSE)
```

Copy OTU column from results file, paste into a new spreadsheet and save as a csv

Prune original OTU table to only include CRT, save to directory 
```{r, cache = TRUE, cache.lazy=FALSE}
speciesList <- tapply(sample_names(av), get_variable(av, "SampleID"), c)
speciesPhyseq <- lapply(speciesList, prune_samples, av)
speciesOTUtable <- lapply(speciesPhyseq,otu_table)
speciesAvg <- lapply(speciesOTUtable,rowMeans)
pooledOTUtable = t(do.call(rbind,speciesAvg)) 
pooledOTUtable = data.frame(OTU=row.names(pooledOTUtable),pooledOTUtable)
avcrt <- read.csv("AV_TP_CRT.csv", header = T)
colnames(avcrt)[1] <- "OTU"
pOTU.crt = merge(pooledOTUtable, avcrt, by.x = "OTU")
 
write.table(pOTU.crt, "av_tp_sid_crt_otu.txt", quote = FALSE, row.names = FALSE, sep = "\t")
```

Convert to biom in qiime

Import CRT biom files, add metadata and taxonomy
```{r}
av_tp_crt_biom <- paste("av_tp_sid_crt_otu.biom", sep = "")
av_tp_crt_biom <- import_biom(av_tp_crt_biom)
av_tp_crt <- merge_phyloseq(av_tp_crt_biom, map, tax_table(av))
```

Make Bray Curtis distance matrices
```{r}
##Time CRT only
av_tp_crt_dist <- distance(av_tp_crt, "bray")

##Whole community + Time CRT
av_dist <- distance(av, "bray")

##Whole community - Time CRT
avtpcrt <- read.csv("AV_TP_CRT.csv", header = T) ###Read in list of CRT that need to be removed
avtpcrt_v <- avcrt[["OTUID"]] ###Make a vector from column of interest
allTaxa <- taxa_names(av) ###Pull out taxa list from phyloseq object
allTaxa <- allTaxa[!(allTaxa %in% avtpcrt_v)] ###Remove CRT from list
av_notpcrt <- prune_taxa(allTaxa, av) ###Prune OTU table to exclude CRT
av_notpcrt_dist <- distance(av_notpcrt, "bray") ###make distance matrix
```

Mantel tests
```{r}
##Whole community + CRT v. Time CRT only
mantel(av_dist, av_tp_crt_dist, method = "pearson")

##Whole community + CRT v. Whole community - Time CRT only
mantel(av_dist, av_notpcrt_dist, method = "pearson")

```

ANOSIM tests
```{r}
##Time CRT
avtp_group = get_variable(av_tp_crt, "TimePoint")
avtp_ano = anosim(av_tp_crt_dist, avtp_group)
avtp_ano$signif
avtp_ano$statistic

##Whole community + CRT
av_group = get_variable(av, "TimePoint")
av_ano = anosim(av_dist, av_group)
av_ano$signif
av_ano$statistic

##Plot results for supplementary figures 4-5
plot(avtp_ano)
plot(av_ano)
```

Outputs from Mantel and ANOSIM entered into a CSV

Make Figure 1
```{r}
##Mantel boxplot
mantel <- read.csv("Mantel_T.csv")
p <- ggplot(mantel, aes(Community, Mantel_r)) + geom_boxplot() + theme_minimal() + geom_point(aes(color = pH, shape = LandUse), size = 5, position = "jitter") + scale_color_gradient(low = "gray0", high = "paleturquoise1") + theme(text = element_text(size=25), axis.title.x = element_blank()) +  scale_x_discrete(labels="CRT v. Whole Commmunity") + labs(y = expression ("Mantel"~R^2)) + scale_shape_discrete(name  ="Land Use")

##ANOSIM boxplots
ano <- read.csv("Anosim_T.csv")
p1 <- ggplot(ano, aes(Community, Anosimr)) + geom_boxplot() + theme_minimal() + geom_point(aes(color = pH, shape = LandUse), size = 5, position = "jitter") + scale_color_gradient(low = "gray0", high = "paleturquoise1") + theme(text = element_text(size=25), axis.title.x = element_blank()) + labs(y = expression ("ANOSIM"~R^2)) + scale_shape_discrete(name ="Land Use")

```

For space-responsive CRT:

Make OTU table that combines site replicates under each sample, across time points 
```{r, cache = TRUE, cache.lazy=FALSE}
speciesList <- tapply(sample_names(mm_r), get_variable(mm_r, "Sample"), c)
speciesPhyseq <- lapply(speciesList, prune_samples, mm_r)
speciesOTUtable <- lapply(speciesPhyseq,otu_table)
speciesAvg <- lapply(speciesOTUtable,rowMeans)
pooledOTUtable = t(do.call(rbind,speciesAvg)) 
pooledOTUtable = data.frame(OTU=row.names(pooledOTUtable),pooledOTUtable)

write.table(pooledOTUtable, "mm_r_sam_otu.txt", quote = FALSE, row.names = FALSE, sep = "\t")
```

Convert to biom in qiime

Import biom file, add metadata and taxonomy
```{r}
mm_r_sam_biom <- paste("mm_r_sam_otu.biom", sep = "")
mm_r_sam_biom <- import_biom(mm_r_sam_biom)
mm_r_sam_map <- paste("MM_All_mapping_BySample_R.txt", sep = "")
mm_r_sam_map <- import_qiime_sample_data(mm_r_sam_map)
mm_r_sam <- merge_phyloseq(mm_r_sam_biom, tax_table(mm_r), mm_r_sam_map)
```

Round counts
```{r}
mm_r_sam_r  = transform_sample_counts(mm_r_sam, round)
```
Plot rank abundance curve (figure s3)
```{r}
library(dplyr)
phyloTemp = transform_sample_counts(mm_r_sam_r, function(x) 1e+02 * x/sum(x))
clusterData = psmelt(phyloTemp)
clusterData = filter(clusterData,Abundance > 0)
clusterAgg = aggregate(Abundance ~ OTU + Phylum,data=clusterData,mean)
clusterAgg = clusterAgg[order(-clusterAgg$Abundance),]
ggplot(clusterAgg,aes(x=reorder(OTU,-Abundance),y=Abundance)) +
    geom_point(aes(color=Phylum),size=3) + 
    theme(axis.ticks = element_blank(), axis.text.x = element_blank()) +
    scale_y_log10()
```

Use rank abundance curve to choose a cutoff for the rare biosphere

Get CRT
```{r}
mm_r_sam_r_dat <- data.frame(otu_table(mm_r_sam_r))
write.table(mm_r_sam_r_dat, "mm_r_sam_r_otu.txt", quote = FALSE,  sep = "\t")

SimpleRareToPrev.f(otu_fp="mm_r_sam_r_otu.txt", abund_thresh=0.0001, abund_thresh_ALL=FALSE, b_thresh=0.90, rdp_lastcol=FALSE)
```

Copy OTU column from results file, paste into a new spreadsheet and save as a csv

Prune OTU table to only include CRT, save to directory 
```{r, cache = TRUE, cache.lazy=FALSE}
speciesList2 <- tapply(sample_names(mm_r), get_variable(mm_r, "SampleID"), c)
speciesPhyseq2 <- lapply(speciesList2, prune_samples, mm_r)
speciesOTUtable2 <- lapply(speciesPhyseq2,otu_table)
speciesAvg2 <- lapply(speciesOTUtable2,rowMeans)
pooledOTUtable2 = t(do.call(rbind,speciesAvg2)) 
pooledOTUtable2 = data.frame(OTU=row.names(pooledOTUtable2),pooledOTUtable2)
mm_r_sam_crt <- read.csv("MM_R_SAM_CRT.csv", header = T)
colnames(mm_r_sam_crt)[1] <- "OTU"
pOTU.crt = merge(pooledOTUtable2, mm_r_sam_crt, by.x = "OTU")

write.table(pOTU.crt, "mm_r_sam_sid_crt.txt", quote = FALSE, row.names = FALSE, sep = "\t")
```

Convert to biom in qiime

Import CRT biom file, add metadata and taxonomy
```{r}
mm_r_sam_crt_biom <- paste("mm_r_sam_sid_crt.biom", sep = "")
mm_r_sam_crt_biom <- import_biom(mm_r_sam_crt_biom)
mm_r_sam_crt <- merge_phyloseq(sample_data(mm_r),mm_r_sam_crt_biom)
mm_r_sam_crt <- merge_phyloseq(mm_r_sam_crt, tax_table(mm_r))
```

Make Bray Curtis distance matrices
```{r}
##Whole community + CRT
mm_r_dist <- distance(mm_r, "bray")

##Space CRT only
mm_r_sam_crt_dist <- distance(mm_r_sam_crt, "bray")

##Whole community - Space CRT
samcrt <- read.csv("MM_R_SAM_CRT.csv", header = T) ###Read in list of CRT that need to be removed
samcrt_v <- samcrt[["OTUID"]] ###Make a vector from column of interest
allTaxa <- taxa_names(mm_r) ###Pull out taxa list from phyloseq object
allTaxa <- allTaxa[!(allTaxa %in% samcrt_v)] ###Remove CRT from list
mm_r_nosamcrt <- prune_taxa(allTaxa, mm_r) ###Prune OTU table to exclude CRT
mm_r_nosamcrt_dist <- distance(mm_r_nosamcrt, "bray") ###make distance matrix
```

Mantel tests
```{r}
##Whole community + CRT v. Space CRT only
mantel(mm_r_dist, mm_r_sam_crt_dist, method = "pearson")

##Whole community + CRT v. Whole community - Space CRT only
mantel(mm_r_dist, mm_r_nosamcrt_dist, method = "pearson")

```

DCA plots showing relationship between CRT, whole community and spatial factors (figure s6)
```{r}
##Make ordinations
mm_r_sam_crt_dca = ordinate(mm_r_sam_crt, "DCA", "bray") ###CRT
mm_r_dca = ordinate(mm_r, "DCA", "bray") ###whole community

##pH/Land Use DCAs
p2 = plot_ordination(mm_r_sam_crt, mm_r_sam_crt_dca, color = "pH", shape = "LandUse") + theme_few() + theme(text = element_text(size=25), axis.text.x = element_text(angle = 90, hjust = 1)) + scale_color_gradient(low = "thistle", high = "skyblue3") + geom_point(size = 5) + scale_shape_discrete(name ="Land Use") + ggtitle("CRT")

p3 = plot_ordination(mm_r, mm_r_dca, color = "pH", shape = "LandUse") + theme_few() + theme(text = element_text(size=25), axis.text.x = element_text(angle = 90, hjust = 1)) + scale_color_gradient(low = "thistle", high = "skyblue3") + geom_point(size = 5) + scale_shape_discrete(name ="Land Use") + ggtitle("Whole Community")

##Soil Order DCAs
p4 = plot_ordination(mm_r_sam_crt, mm_r_sam_crt_dca, color = "SoilOrder") + theme_few() + theme(text = element_text(size=25), axis.text.x = element_text(angle = 90, hjust = 1)) + geom_point(size = 5) + scale_color_discrete(name = "Soil Order") + ggtitle("CRT") 

p5 = plot_ordination(mm_r, mm_r_dca, color = "SoilOrder") + theme_few() + theme(text = element_text(size=25), axis.text.x = element_text(angle = 90, hjust = 1)) + geom_point(size = 5) + scale_color_discrete(name = "Soil Order") + ggtitle("Whole Community") 
```

ANOSIM tests against land use and soil order (figure s7, table s2)
```{r}
##CRT v. Land Use
mm_r_crtlu_group = get_variable(mm_r_sam_crt, "LandUse")
mm_r_crtlu_ano = anosim(mm_r_sam_crt_dist, mm_r_crtlu_group)
mm_r_crtlu_ano$signif
mm_r_crtlu_ano$statistic

##Whole community v. Land Use
mm_r_lu_group = get_variable(mm_r, "LandUse")
mm_r_lu_ano = anosim(mm_r_dist, mm_r_lu_group)
mm_r_lu_ano$signif
mm_r_lu_ano$statistic

##CRT v. Soil Order
mm_r_crtso_group = get_variable(mm_r_sam_crt, "SoilOrder")
mm_r_crtso_ano = anosim(mm_r_sam_crt_dist, mm_r_crtso_group)
mm_r_crtso_ano$signif
mm_r_crtso_ano$statistic

##Whole community v. Soil Order
mm_r_so_group = get_variable(mm_r_, "SoilOrder")
mm_r_so_ano = anosim(mm_r_dist, mm_r_so_group)
mm_r_so_ano$signif
mm_r_so_ano$statistic

##Plot results for supplementary figure 7
plot(mm_r_crtlu_ano)
plot(mm_r_lu_ano)
plot(mm_r_crtso_ano)
plot(mm_r_so_ano)
```

Mantel tests against pH (table s2)
```{r}
#Make pH distance matrix, will be same for both CRT and whole community
mm_r_sd <- data.frame(mm_r_map)
mm_r_ph_dist <- dist(mm_r_sd$pH, method = "euclidean")
mm_r_ph_dist <- as.matrix(mm_r_ph_dist)

#CRT v. pH
mantel(mm_r_sam_crt_dist, mm_r_ph_dist, method = "pearson")

#Whole community v. pH
mantel(mm_r_dist, mm_r_ph_dist, method = "pearson")
```

Correlations between individual CRTs and spatiotemporal factors
```{r}
library(ALDEx2)

##Time
av_tp_crt_otu <- data.frame(otu_table(av_tp_crt)) ###Make OTU table into data frame
av_meta <- data.frame(sample_data(av)) ###Make sample data into data frame

avtpcrt.aldex.clr <- aldex.clr(av_tp_crt_otu, mc.samples = 128) ###generate monte carlo samples, log transform data 
avtpcrt.time.aldex.glm <- aldex.glm(avtpcrt.aldex.clr, av_meta$TimePoint) ###generate correlations between OTUs and time

##Space
mm_r_sam_crt_otu <- data.frame(otu_table(mm_r_sam_crt)) ###Make OTU table into data frame
meta <- data.frame(map) ###Make sample data into data frame

mmrsam.aldex.clr <- aldex.clr(mm_r_sam_crt_otu, mc.samples = 128) ###generate monte carlo samples, log transform data 
mmrsam.lu.aldex.glm <- aldex.glm(mmrsam.aldex.clr, meta$LandUse) ###generate correlations between OTUs and land use
mmrsam.so.aldex.glm <- aldex.glm(mmrsam.aldex.clr, meta$SoilOrder) ###generate correlations between OTUs and soil order
mmrsam.pH.aldex.corr <- aldex.corr(mmrsam.aldex.clr, meta$pH) ###generate correlations between OTUs and pH
```

Link OTUs to sapatiotemporal factors to plot
```{r}
##Convert to relative abundance of a sample
mm_r_sam_crt_r  = transform_sample_counts(mm_r_sam_crt, function(x) x / sum(x))

##Average samples across Land Use, convert to data frame, attach taxonomy and prune to only include significantly correlated CRT from ALDEx2 output
speciesList1 <- tapply(sample_names(mm_r_sam_crt_r), get_variable(mm_r_sam_crt_r, "LandUse"), c)
speciesPhyseq1 <- lapply(speciesList1, prune_samples, mm_r_sam_crt_r)
speciesOTUtable1 <- lapply(speciesPhyseq1,otu_table)
speciesAvg1 <- lapply(speciesOTUtable1,rowMeans)
pooledOTUtable1 = t(do.call(rbind,speciesAvg1)) 
pooledOTUtable1 = data.frame(OTU=row.names(pooledOTUtable1),pooledOTUtable1)
TT = tax_table(mm_r_sam_crt_r)
TT = TT[, which(apply(!apply(TT, 2, is.na), 2, any))]
tdf = data.frame(TT, OTU = taxa_names(mm_r_sam_crt_r))
pOTUtax1 = merge(pooledOTUtable1, tdf, by.x = "OTU")
sigluOTUs = read.csv("LU_Aldex.csv")
pOTUtax1 = merge(pOTUtax1, sigluOTUs, by= "OTU")
library(reshape2)
pOTUtax1.m <- melt(pOTUtax1)
write.csv(pOTUtax1.m, "LU_Aldex_melt.csv")

##Repeat above for soil order and pH

##Merge site-level phyloseq objects
tp_crt <- merge(av_tp_crt, back_tp_crt...)
##Convert to relative abundance of a sample
tp_crt_r  = transform_sample_counts(tp_crt, function(x) x / sum(x))

####Average samples across Time Point, convert to data frame, attach taxonomy and prune to only include significantly correlated CRT from ALDEx2 output-used most significant OTU from each site, as there were too many correalted OTUs to plot
speciesList2 <- tapply(sample_names(tp_crt_r), get_variable(tp_crt_r, "TimePoint"), c)
speciesPhyseq2 <- lapply(speciesList2, prune_samples, tp_crt_r)
speciesOTUtable2 <- lapply(speciesPhyseq2,otu_table)
speciesAvg2 <- lapply(speciesOTUtable2,rowMeans)
pooledOTUtable2 = t(do.call(rbind,speciesAvg2)) 
pooledOTUtable2 = data.frame(OTU=row.names(pooledOTUtable2),pooledOTUtable2)
TT2 = tax_table(tp_crt_r)
TT2 = TT2[, which(apply(!apply(TT2, 2, is.na), 2, any))]
tdf2 = data.frame(TT2, OTU = taxa_names(tp_crt_r))
pOTUtax2 = merge(pooledOTUtable2, tdf2, by.x = "OTU")
sigtpOTUs = read.csv("TP_Aldex.csv")
pOTUtax2 = merge(pOTUtax2, sigtpOTUs, by= "OTU")
library(reshape2)
pOTUtax2.m <- melt(pOTUtax2)
write.csv(pOTUtax2.m, "TP_Aldex_melt.csv")
```

Add column to csv file that has the lowest available taxonomic classification

Make Figure 2
```{r}
pOTUtaxm.m <- read.csv("LU_Aldex_melt.csv")
name_order <- c("Dairy", "Sheep+Beef", "HighCountry")
pOTUtaxm.m$LandUse <- factor(pOTUtaxm.m$LandUse, levels=name_order)
g <- ggplot(subset(pOTUtaxm.m, Abundance >0), aes(x=LandUse, y=Taxon, color=Phylum))
g <- g + geom_point(aes(size=Abundance)) + scale_color_manual(values = c("light coral", "springgreen3", "cyan3", "mediumpurple2")) + labs(size = "Relative Abundance") + scale_size(range = c(3, 13))
g <- g + theme_minimal()
g <- g + theme(text = element_text(size=25)) + scale_x_discrete(name= "Land Use", labels= c("Dairy", "Sheep + Beef", "High Country"))
g <- g + guides(color = guide_legend(override.aes = list(size=5)), size = guide_legend(order=1))

##Repeat above for soil order, time point and pH 
```
